Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-21 aistats.tex
Complete proof of self-concordance

\paragraph{Log-partition is Self-Concordant}
Adopting the notation of definition~\ref{def:self-concordance}, let us set $w=(\nat_1 \ 0)^\top$ and $v=(0 \ -1)^\top$.
Then 
\alignn{g(t) := 2A(w+tv) = \frac{a}{t} - \log(t)}
is well defined for $t>0$ and $a=\frac{\nat_1^2}{2} \geq 0$.
The logarithm is self-concordant for the same reason that the entropy is, but the inverse function $\inv{t}$ is not \citep[example 4.1.1.5]{nesterov2003introductory}.
Linear combinations with positive coefficients  preserves self-concordance \citep[theorem 4.1.1]{nesterov2003introductory} -- e.g. self-concordance is a cone, so it seems unlikely that this sum of a self-concordant function and a not self-concordant function would be self-concordant. 
But let us verify it to be sure. The derivatives of $g$ are
\alignn{
g'(t) &= -\frac{a}{t^2} - \frac{1}{t} \; ,\\
g''(t) &= \frac{2a}{t^3} + \inv{t^2} = \frac{1}{t^2}\paren{\frac{2a}{t} + 1} \; , \\
g'''(t) &= -\frac{6a}{t^4} - \frac{2}{t^3} = \frac{-2}{t^3}\paren{\frac{3a}{t} + 1} \;.
}
Then the definition of self-concordance reads
\alignn{
\abs{g'''(t)} &\leq 2 g''(t)^{\frac{3}{2}} \\
\iff \frac{3a}{t} + 1 &\leq \paren{\frac{2a}{t} + 1}^{\frac{3}{2}} \; .
}
Writing $y = \frac{a}{t} \geq 0$, and taking the square of this expression yields
\alignn{
 (3y + 1)^2 &\leq (2y + 1)^3 \\
 \iff 9 y ^2 + 6y + 1 &\leq 8 y^3  + 12 y^2 + 6y +1 \\
\iff 0 &\leq y^2(8y + 3)
}
which is true since $y\geq 0$.
The log-partition is self concordant.

>>>>

\paragraph{Log-partition is Self-Concordant}
Adopting the notation of definition~\ref{def:self-concordance}, let us set $w=(\nat_1 \ 0)^\top$ and $v=(\alpha \ \ -1)^\top$. 
For varying $\nat_1, \alpha$, this covers the whole range of possible straight lines in the 2D plane, except lines of constant $\nat_2$ (horizontal).
But on lines with $\nat_2=\cst$,   the log-partition function $A$ is simply a quadratic, so it is self-concordant.
Then, up to constants,
\alignn{
g(t) := 2A(w+tv) 
= \frac{(\nat_1 + t \alpha)^2}{2 t} - \log(t) + \cst
= \frac{a}{t}  + b t - \log(t) + \cst
}
where $a=\frac{\nat_1^2}{2} \geq 0$, and $b = \frac{\alpha^2}{2}$.
$g$ is well defined for $t>0$.
The logarithm is self-concordant for the same reason that the entropy is, but the inverse function $\inv{t}$ is not \citep[example 4.1.1.5]{nesterov2003introductory}.
Linear combinations with positive coefficients  preserves self-concordance \citep[theorem 4.1.1]{nesterov2003introductory} -- e.g. self-concordance is a cone, so it seems unlikely that this sum of a self-concordant function and a not self-concordant function would be self-concordant. 
So let us prove it.
The derivatives of $g$ are
\alignn{
g'(t) &= -\frac{a}{t^2} - \frac{1}{t} + b \; ,\\
g''(t) &= \frac{2a}{t^3} + \inv{t^2} = \frac{1}{t^2}\paren{\frac{2a}{t} + 1} \; , \\
g'''(t) &= -\frac{6a}{t^4} - \frac{2}{t^3} = \frac{-2}{t^3}\paren{\frac{3a}{t} + 1} \;.
}
Then the definition of self-concordance reads
\alignn{
\abs{g'''(t)} &\leq 2 g''(t)^{\frac{3}{2}} \\
\iff \frac{3a}{t} + 1 &\leq \paren{\frac{2a}{t} + 1}^{\frac{3}{2}} \; .
}
Writing $y = \frac{a}{t} \geq 0$, and taking the square of this expression yields
\alignn{
 (3y + 1)^2 &\leq (2y + 1)^3 \\
 \iff 9 y ^2 + 6y + 1 &\leq 8 y^3  + 12 y^2 + 6y +1 \\
\iff 0 &\leq y^2(8y + 3)
}
which is true since $y\geq 0$.
The log-partition is self concordant.
